The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-03 manuscript.Rmd
Text improvements and corrections, spellcheck
Co-authored-by: Dirk Eddelbuettel <edd@debian.org>

  This articles surveys downstream projects that build upon Rocker images and presents the current state of R packages for managing Docker images and controlling containers.

@@

  The variety of applications demonstrates the power of Rocker specifically and containerisation in general.

@@

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Also though, coinciding with the rise of R has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specifically for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
We conclude with by reflecting on the landscape of packages and applications and point out future directions of development.

@@

Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020},
makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018} because, while a container typically runs one process, virtualisation may run whole operating systems at a larger footprint.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
The success of Docker has even led to standardisation and industry collaboration \citep{oci_open_2019}.

@@

Docker is not the only containerisation software.
An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
Singularity can run Docker images.
Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

Interfacing with the Docker daemon is typically done through the [Docker Command Line Interface](https://docs.docker.com/engine/reference/commandline/cli/) (Docker CLI).
However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
A number of first-order R packages provide an interface to the Docker CLI, allowing for the interaction with the Docker CLI from an R console, e.g., to be automated.

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or in the cloud.
As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.

@@

tab <-
  full_join(
  functionalities %>% crossing(package = package_functionalities$package),
  package_functionalities %>% unnest(id) %>% mutate(check = "\\checkmark"),#\u2714 # \\checkmark
  by = c("id", "package")
) %>%
  mutate(check = ifelse(is.na(check), "", check)) %>%
  spread(package, check) %>%
  slice(match(functionalities$id, id)) %>%
  select(-id)

# insert table
requireNamespace("kableExtra")
knitr::kable(tab, format = "latex", escape = FALSE) %>%
  kableExtra::row_spec(0, angle = -90)
```

@@

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
The remaineder of this section presents exemple Docker images and image stacks featuring R intended for data science.

@@

Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.

@@

First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
The container does not affect their regular system.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmisation during software evaluation.
\citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

Fourth, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally; this can be a challenge for some users.
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one \citep{r-hub_docs_2019}.
R-hub makes it easy to ensure that no errors occur, but fixing errors still often warrants a local setup, e.g., using the image `rocker/r-devel`, as is testing packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but they lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems with which they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly seperates test environment, test code, and current working copy of the code.
\label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but it requires users to manually manage the containers.

\label{ci}
Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018} also saved time with Travis&nbsp;CI by re-using the testing image as the basis for an image intended for publication on Docker&nbsp;Hub.
[`r-ci`](https://github.com/ColinFay/r-ci) is, in turn, used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image and control code, and the whole set of tests is run inside a container.
The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
Especially for long-running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
Containers can also simplify the integration of R software into larger, multi-language CI pipelines.
Furthermore, with each change, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository), not because of concern about time, but to control the environment used on a CI server.
This gives, on the one hand, easy access after every update of the R Markdown source code and, on the other hand a second controlled environment to make sure that the article renders successfully and correctly.

@@

First, it is possible to **offload complex processing to a server** or clouds and also to execute proccesses in parallel to speed up or to serve many users.

@@

It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.

@@

Futhermore, \CRANpkg{drake} workflwos have been shown to use \CRANpkg{future} package's function `makeClusterPSOCK()` for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.

@@

First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications \citep{cran_shiny}.
_ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
The tool is built on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executing long-running R jobs or interactive R sessions.
The started containers can run on a regular Docker host but also in clusters.
Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy can be achieved, e.g., via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html).
A pipeline can include building and checking R packages and Shiny apps.
After the code has passed the checks, Docker images are built and pushed to GitLab's container registry.
The pipeline finishes with triggering a webhook on the server, where the
deployment script is executed.
The script updates configurations, pulls the new Docker images, and restarts the services.

@@

Independent integrations of R for different cloud providers lead to repeated efforts and code fragmentation.
To mitigate these problems and to avoid vendor lock-in motivated the [OpenFaaS project](https://www.openfaas.com/).
OpenFaas facilitates the deployment of functions and microservices to Kubernetes or Docker Swarm.
It is language agnostic and provides auto-scaling, metrics, and an API gateway.
Reduced boilerplate code is achieved via templates.
Template for R\footnote{See OpenFaaS R templates at \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}.}
are provided based on Rocker's Debian and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Alpine images.
The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver.
The watchdog marshals an HTTP request and invokes the actual application.
The R session uses \CRANpkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.

The prevelance of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercializable features.
One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are publicly available.

>>>>

  This articles surveys downstream projects that build upon the Rocker Project images and presents the current state of R packages for managing Docker images and controlling containers.

@@

  The variety of applications demonstrates the power of the Rocker Project specifically and containerisation in general.

@@

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
During the last few years, and coinciding with the rise of R, has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents what we may call the _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specifically for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
We conclude by reflecting on the landscape of packages and applications and point out future directions of development.

@@

Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020},
makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}, because virtualisation may emulate a whole operating system, a container typically runs only one process.
The single process together with sharing the host's kernel results in a reduced footprint and faster start times.
While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
The success of Docker has even paved the way for industry collaboration and standardisation \citep{oci_open_2019}.

@@

Docker is not the only containerisation software.
An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
Singularity can run Docker images.
Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
To date, no comparable image stack to the Rocker Project's images exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

Users interact with the Docker daemon typically through the [Docker Command Line Interface](https://docs.docker.com/engine/reference/commandline/cli/) (Docker CLI).
However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility because of manual steps.
A number of first-order R packages provide an interface to the Docker CLI, allowing for the interaction with the Docker CLI from an R console.
Table&nbsp;\ref{tab:clients} gives an overview of packages with client functionality, each of which provides functions for interacting with the Docker daemon.
The packages focus on different aspects and support different stages of a container's life cycle.
As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.

@@

tab <-
  full_join(
  functionalities %>% crossing(package = package_functionalities$package),
  package_functionalities %>% unnest(id) %>% mutate(check = "\\checkmark"),#\u2714 # \\checkmark # ✓
  by = c("id", "package")
) %>%
  mutate(check = ifelse(is.na(check), "", check)) %>%
  spread(package, check) %>%
  slice(match(functionalities$id, id)) %>%
  select(-id)

# insert table, position caption below as described in https://github.com/haozhu233/kableExtra/issues/221#issuecomment-400473530
library("kableExtra")
library("xtable")
#knitr::kable(tab,
#             format = "latex",
#             caption = "R packages with Docker client functionality.",
#             label = "clients",
#             escape = FALSE) %>%
#  kableExtra::row_spec(0, angle = -90)
xtable2kable <- function(x) {
  label(x) <- "tab:clients"
  out <- capture.output(print(x, 
                              table.placement = NULL,
                              booktabs = TRUE,
                              caption.placement = "bottom",
                              include.rownames = FALSE,
                              hline.after = c(-1:nrow(x)),
                              # next analoguosu to kable's "escape = FALSE", see https://stackoverflow.com/a/42355065/261210
                              sanitize.text.function = function (x) x
                              ))[-(1:2)]
  out <- paste(out, collapse = "\n")
  structure(out, format = "latex", class = "knitr_kable")
}

xtable(tab, caption = "R packages with Docker client functionality.") %>%
  xtable2kable() %>%
  kableExtra::row_spec(0, angle = -90)
```

@@

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
The remainder of this section presents example Docker images and image stacks featuring R intended for data science.

@@

Both approaches enable the `apt` package manager \citep{wikipedia_contributors_apt_2020} as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker images which rely on plain `Dockerfile`s.

@@

First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
The container does not affect their regular system.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for customisation during software evaluation.
\citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occurring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accommodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

Fourth, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally; this can be a challenge for some users.
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one \citep{r-hub_docs_2019}.
R-hub makes it easy to ensure that no errors occur, but fixing errors still often warrants a local setup, e.g., using the image `rocker/r-devel`, as is testing packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but they lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems with which they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly separates test environment, test code, and current working copy of the code.
\label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but it requires users to manually manage the containers.

\label{ci}
Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018} also saved time with Travis&nbsp;CI by re-using the testing image as the basis for an image intended for publication on Docker&nbsp;Hub.
[`r-ci`](https://github.com/ColinFay/r-ci) is, in turn, used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image and control code, and the whole set of tests is run inside a container.
The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
Especially for long-running tests or complex system dependencies, these approaches to separate installation of build dependencies with code testing streamline the development process.
Containers can also simplify the integration of R software into larger, multi-language CI pipelines.
Furthermore, with each change, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository), not because of concern about time, but to control the environment used on a CI server.
This gives, on the one hand, easy access after every update of the R Markdown source code and, on the other hand a second controlled environment to make sure that the article renders successfully and correctly.

@@

First, it is possible to **offload complex processing to a server** or clouds and also to execute processes in parallel to speed up or to serve many users.

@@

It features implicit parallel computing and automated detection of the parts of the work that actually needs to be re-executed.

@@

Furthermore, \CRANpkg{drake} workflows have been shown to use \CRANpkg{future} package's function `makeClusterPSOCK()` for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.

@@

First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications \citep{cran_shiny}.
_ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarios with multiple apps.
ShinyProxy itself is written in Java to accommodate corporate requirements and may itself run in a container for stability and availability.
The tool is built on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executing long-running R jobs or interactive R sessions.
The started containers can run on a regular Docker host but also in clusters.
Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy can be achieved, e.g., via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html).
A pipeline can include building and checking R packages and Shiny apps.
After the code has passed the checks, Docker images are built and pushed to GitLab's container registry.
The pipeline finishes with triggering a webhook on the server, where the
deployment script is executed.
The script updates configurations, pulls the new Docker images, and restarts the services.

@@

Independent integrations of R for different cloud providers lead to repeated efforts and code fragmentation.
To mitigate these problems and to avoid vendor lock-in motivated the [OpenFaaS project](https://www.openfaas.com/).
OpenFaas facilitates the deployment of functions and microservices to Kubernetes or Docker Swarm.
It is language agnostic and provides auto-scaling, metrics, and an API gateway.
Reduced boilerplate code is achieved via templates.
Template for R\footnote{See OpenFaaS R templates at \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}.}
are provided based on Rocker's Debian and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Alpine images.
The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang web server.
The watchdog marshals an HTTP request and invokes the actual application.
The R session uses \CRANpkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.

The prevalence of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercialisation features.
One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commercial products are publicly available.
